---
title: "Pedestrian Movement Patterns in Melbourne"
subtitle: "ETC5513 Assignment 3"
author:
  - "Aarushi"
  - "Abhigna"
  - "Aditi Srinivas"
  - "Xijun Wei"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    smaller: true
    scrollable: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup-slides
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(kableExtra)

pedestrian_data <- read_csv(
  "data/pedestrian-counting-system-monthly-counts-per-hour-2.csv"
)

pedestrian_summer <- pedestrian_data |>
  mutate(Sensing_Date = as.Date(Sensing_Date)) |>
  filter(
    Sensing_Date >= as.Date("2025-12-01"),
    Sensing_Date <= as.Date("2026-02-28")
  )
```

## Problem Introduction

- Melbourne CBD generates millions of pedestrian movements daily
- Understanding **where** and **when** people move is critical for city planning
- Summer brings unique patterns driven by tourism, retail and events
- **Research question:** How do pedestrian volumes vary by location and time during the 2025-26 summer?

---

## Dataset Description

- Source: City of Melbourne Open Data Portal
- Pedestrian Counting System - automated sensors across the CBD
- Filtered to: 1 December 2025 to 28 February 2026
- Key variables:
  - `Total_of_Directions` - hourly pedestrian count per sensor
  - `Sensor_Name` - sensor location
  - `Sensing_Date` and `HourDay` - date and hour of observation

---

## Methods

- Data filtered to summer period using R and tidyverse
- Descriptive statistics calculated: total, mean and maximum counts
- Top sensor locations identified by average hourly count
- Daily trends examined for the five busiest sensors
- All analysis fully reproducible using renv and Quarto

---

## Results - Daily Trends

- Pedestrian activity **highest in December**, declines from January

```{r}
#| fig-width: 9
#| fig-height: 4

invisible(Sys.setlocale("LC_TIME", "C"))

top_5 <- pedestrian_summer |>
  group_by(Sensor_Name) |>
  summarise(
    Average_Count = mean(Total_of_Directions, na.rm = TRUE)
  ) |>
  arrange(desc(Average_Count)) |>
  slice_head(n = 5)

pedestrian_summer |>
  filter(Sensor_Name %in% top_5$Sensor_Name) |>
  group_by(Sensing_Date, Sensor_Name) |>
  summarise(
    mean_count = mean(Total_of_Directions),
    .groups = "drop"
  ) |>
  ggplot(aes(Sensing_Date, mean_count)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.8) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  facet_wrap(~Sensor_Name, ncol = 3) +
  labs(x = "Date", y = "Average Count") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  )
```

---

## Results - Top 10 Sensor Locations

- **Swa123_T** busiest sensor (3.63M pedestrians, 4.5% of total traffic)
- Most sensors peak **at 5 PM**; two peak at **1 PM**

```{r}
#| fig-width: 8
#| fig-height: 4

location_summary <- pedestrian_summer |>
  group_by(Sensor_Name) |>
  summarise(
    Total_Count = sum(Total_of_Directions),
    Mean_Count = round(mean(Total_of_Directions), 0),
    Max_Count = max(Total_of_Directions)
  ) |>
  mutate(Pct_Total_Traffic = round(Total_Count / sum(Total_Count) * 100, 2))

peak_hour <- pedestrian_summer |>
  group_by(Sensor_Name, HourDay) |>
  summarise(
    Hourly_Mean = round(mean(Total_of_Directions), 0),
    .groups = "drop"
  ) |>
  group_by(Sensor_Name) |>
  slice_max(Hourly_Mean, n = 1, with_ties = FALSE) |>
  rename(Peak_Hour = HourDay, Peak_Hour_Mean = Hourly_Mean)

location_summary |>
  left_join(peak_hour, by = "Sensor_Name") |>
  arrange(desc(Total_Count)) |>
  slice(1:10) |>
  mutate(Rank = row_number()) |>
  select(Rank, Sensor_Name, Total_Count, Pct_Total_Traffic,
         Mean_Count, Max_Count, Peak_Hour, Peak_Hour_Mean) |>
  knitr::kable(
    digits = 0,
    format.args = list(big.mark = ",")
  ) |>
  kableExtra::kable_styling(font_size = 14, full_width = FALSE)
```

---

## Conclusions

- Pedestrian movement in Melbourne CBD is not random or evenly distributed
- Activity is shaped by two key factors: **place** and **time**
- The CBD functions as a collection of distinct activity zones, each with its own rhythm
- Effective pedestrian planning must be targeted and location-specific, not one-size-fits-all

---

## Recommendations

- Prioritise pedestrian planning in high-volume areas such as **Swanston Street, Southbank and Elizabeth Street**
- Improve pedestrian flow through clearer signage, better footpath design and crowd management during peak periods
- Align public transport and cleaning services with the busiest times, especially around the **17:00 peak**
- Use pedestrian count data to support event planning and manage pressure during busy holiday periods
- Future analysis should compare summer with other seasons and include factors such as weather, events and public holidays
